---
title: 'Lab 3: hierarchical N-mixture models'
author: "Ben Goldstein and Krishna Pacifici"
date: "Fall 2024"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

# Install/load the necessary packages and 
tryCatch(library(devtools), 
         error = function(e) {
           install.packages("devtools")
           library(devtools)
         })
tryCatch(library(EcoData), 
         error = function(e) {
           devtools::install_github(repo = "TheoreticalEcology/EcoData", 
                                    dependencies = T, build_vignettes = T)
           library(EcoData)
         })
tryCatch(library(tidyverse), 
         error = function(e) {
           install.packages("tidyverse")
           library(tidyverse)
         })
tryCatch(library(nimble), 
         error = function(e) {
           install.packages("nimble")
           library(nimble)
         })

```

Topics:

 - Estimate an N-mixture model
 - Include a categorical fixed effect
 - Include a random effect
 - Predict from the model using derived quantities


#### Estimating the abundance of pileated woodpeckers

We're estimating the abundance of pileated woodpeckers on some land that
we manage. We have two management units. We went to 16 sites in each management unit 4 times each and counted how many woodpeckers we saw at each site.

At each site, we document the elevation and the habitat type (one of two levels.) 
During each visit, we also document the temperature. Sites are themselves nested 
within management units, which could introduce some nonindependence.

```{r}
dat <- read_csv("https://raw.githubusercontent.com/dochvam/AHM_Labs_2024/main/3_Abundance/woodpecker_data.csv")

summary(dat)
```

We want to estimate an N-mixture model to understand how the abundance of woodpeckers varies with elevation and habitat. We also want to take into account nonindependence between sites nested in four different management units and we want to account for variation in detection probability due to temperature and habitat type.

The N-mixture model we want to fit is defined by the following equations:

$$y_{ij} \sim \text{Binom}(N_i, p_{ij})$$
$$N \sim \text{Pois}(\lambda_i)$$
$$\text{logit}(p_{ij}) = \alpha_0 + \alpha_1 \times \text{Temp}_{ij} + \alpha_2 \times \text{Habitat}_{i}$$
$$\text{log}(\lambda_{i}) = \beta_0 + \beta_1 \times \text{Elev}_{i} + \beta_2 \times \text{Habitat}_{i}$$


#### Define and estimate the model

Write the nimbleCode to estimate this model.

```{r}

nmix_code <- nimbleCode({
  # Model code goes here:
  for (i in 1:nsite) {
    for (j in 1:nvisit) {
      logit(p[i,j]) <- alpha0 + alpha1 * temp[i, j] + alpha2 * habitat[i]
      
      y[i,j] ~ dbinom(size = N[i], prob = p[i,j])
    }
    
    log(lambda[i]) <- beta0 + beta1 * elevation[i] + beta2 * habitat[i]
    
    N[i] ~ dpois(lambda[i])
  }
  
  # Priors go here:
  alpha0 ~ dnorm(0, sd = 5)
  alpha1 ~ dnorm(0, sd = 5)
  alpha2 ~ dnorm(0, sd = 5)
  beta0  ~ dnorm(0, sd = 5)
  beta1  ~ dnorm(0, sd = 5)
  beta2  ~ dnorm(0, sd = 5)
})

```

Preprocess data

```{r}
# Preprocess y and temp data to make them wide-format.

y_wide <- dat %>% 
  mutate(visit_ID = paste0("V", visit_ID)) %>% 
  select(site_ID, visit_ID, observed_count) %>% 
  pivot_wider(names_from = visit_ID, values_from = observed_count) %>% 
  select(-site_ID) %>% 
  as.matrix()

temp_wide <- dat %>% 
  mutate(visit_ID = paste0("V", visit_ID)) %>% 
  select(site_ID, visit_ID, temp_C) %>% 
  pivot_wider(names_from = visit_ID, values_from = temp_C) %>% 
  select(-site_ID) %>% 
  as.matrix()

# Scale numeric covariates

temp_wide[] <- as.numeric(scale(temp_wide[]))
elevation <- as.numeric(scale(dat$elevation))

# Convert two-level factor into a binary variable

habitat <- as.numeric(dat$habitat == "mixed_forest_developed")

```



Build the model. 

```{r}
mod <- nimbleModel(
  nmix_code,
  constants = list(
    nsite = length(unique(dat$site_ID)),
    nvisit = length(unique(dat$visit_ID)),
    elevation = elevation,
    habitat = habitat,
    temp = temp_wide
  ),
  data = list(
    y = y_wide
  ),
  inits = list( 
    alpha0 = 0, alpha1 = 0, alpha2 = 0,
    beta0 = 0, beta1 = 0, beta2 = 0,
    N = rep(max(y_wide) + 5, length(unique(dat$site_ID)))
  )
)
```


Use the NIMBLE workflow to run an MCMC algorithm and obtain posterior samples.

```{r}
cmod <- compileNimble(mod)
mcmc <- buildMCMC(mod)
cMCMC <- compileNimble(mcmc)

samples <- runMCMC(cMCMC, niter = 5000, nburnin = 1000, nchains = 2, thin = 1)

summary1 <- MCMCvis::MCMCsummary(samples)

summary1
```

We should now have estimates of our $\alpha$ and $\beta$ parameters.


#### Add random effects

In the previous model, we ignored potential non-independence arising from the fact that our data are grouped by management units. There could be random effects on either detection or abundance.

In the following code chunk, replicate the previous model, but now include a random effect of management unit on abundance.

```{r}
nmix_code_wranef <- nimbleCode({
  # Model code goes here:
  for (i in 1:nsite) {
    for (j in 1:nvisit) {
      logit(p[i,j]) <- alpha0 + alpha1 * temp[i, j] + alpha2 * habitat[i]
      
      y[i,j] ~ dbinom(size = N[i], prob = p[i,j])
    }
    
    log(lambda[i]) <- beta0 + beta1 * elevation[i] + beta2 * habitat[i] +
      ranef_mgmt[unit[i]]
    
    N[i] ~ dpois(lambda[i])
  }
  
  # Priors go here:
  alpha0 ~ dnorm(0, sd = 5)
  alpha1 ~ dnorm(0, sd = 5)
  alpha2 ~ dnorm(0, sd = 5)
  beta0  ~ dnorm(0, sd = 5)
  beta1  ~ dnorm(0, sd = 5)
  beta2  ~ dnorm(0, sd = 5)
  
  # Random effect priors
  ranef_sd ~ dunif(0, 5)
  for (i in 1:nunit) {
    ranef_mgmt[i] ~ dnorm(0, sd = ranef_sd)
  }
})

mod <- nimbleModel(
  nmix_code_wranef,
  constants = list(
    nsite = length(unique(dat$site_ID)),
    nvisit = length(unique(dat$visit_ID)),
    nunit = length(unique(dat$management_unit)),
    unit = dat$management_unit,
    elevation = elevation,
    habitat = habitat,
    temp = temp_wide
  ),
  data = list(
    y = y_wide
  ),
  inits = list( 
    alpha0 = 0, alpha1 = 0, alpha2 = 0,
    beta0 = 0, beta1 = 0, beta2 = 0,
    ranef_sd = 1, 
    ranef_mgmt = rnorm(length(unique(dat$management_unit)), 0, sd = 0.1),
    N = rep(max(y_wide) + 5, length(unique(dat$site_ID)))
  )
)

cmod <- compileNimble(mod)
mcmc <- buildMCMC(mod)
cMCMC <- compileNimble(mcmc)

# Mixing is much slower! Need lots of iterations
samples <- runMCMC(cMCMC, niter = 50000, nburnin = 10000, nchains = 4, thin = 1)

summary2 <- MCMCvis::MCMCsummary(samples)

summary2

```



#### Derive a predicted abundance 

Now that we have a satisfactory model, we want to use it to make predictions of mean abundance under different conditions. We can do this by making

We can create a dataset that contains hypothetical input data for a number of sites we can imagine predicting abundance at.

```{r}
# use expand.grid to get all combinations of these different variables
prediction_sites <- data.frame(expand.grid(
  elevation = 50:100,
  habitat = c("forest", "mixed_forest_developed"),
  management_unit = 1:4
))

head(prediction_sites)
```

Define, build, and estimate a new nimbleModel to predict the mean abundance ($\lambda_i$) at each of these hypothetical sites.

```{r}

# Need to scale the prediction data the same way!

elevation_derived <- (prediction_sites$elevation - mean(dat$elevation)) / 
                        sd(dat$elevation)
habitat_derived <- as.numeric(prediction_sites$habitat == "mixed_forest_developed")


nmix_code_wderived <- nimbleCode({
  # Model code goes here:
  for (i in 1:nsite) {
    for (j in 1:nvisit) {
      logit(p[i,j]) <- alpha0 + alpha1 * temp[i, j] + alpha2 * habitat[i]
      
      y[i,j] ~ dbinom(size = N[i], prob = p[i,j])
    }
    
    log(lambda[i]) <- beta0 + beta1 * elevation[i] + beta2 * habitat[i] +
      ranef_mgmt[unit[i]]
    
    N[i] ~ dpois(lambda[i])
  }
  
  for (i in 1:nderived) {
    log(lambda_derived[i]) <- beta0 + beta1 * elevation_derived[i] + 
                              beta2 * habitat_derived[i] +
                              ranef_mgmt[unit_derived[i]]
  }
  
  # Priors go here:
  alpha0 ~ dnorm(0, sd = 5)
  alpha1 ~ dnorm(0, sd = 5)
  alpha2 ~ dnorm(0, sd = 5)
  beta0  ~ dnorm(0, sd = 5)
  beta1  ~ dnorm(0, sd = 5)
  beta2  ~ dnorm(0, sd = 5)
  
  # Random effect priors
  ranef_sd ~ dunif(0, 5)
  for (i in 1:nunit) {
    ranef_mgmt[i] ~ dnorm(0, sd = ranef_sd)
  }
})

mod <- nimbleModel(
  nmix_code_wderived,
  constants = list(
    nsite = length(unique(dat$site_ID)),
    nvisit = length(unique(dat$visit_ID)),
    nunit = length(unique(dat$management_unit)),
    unit = dat$management_unit,
    elevation = elevation,
    habitat = habitat,
    temp = temp_wide,
    elevation_derived = elevation_derived,
    habitat_derived = habitat_derived,
    unit_derived = prediction_sites$management_unit,
    nderived = nrow(prediction_sites)
  ),
  data = list(
    y = y_wide
  ),
  inits = list( 
    alpha0 = 0, alpha1 = 0, alpha2 = 0,
    beta0 = 0, beta1 = 0, beta2 = 0,
    ranef_sd = 1, 
    ranef_mgmt = rnorm(length(unique(dat$management_unit)), 0, sd = 0.1),
    N = rep(max(y_wide) + 5, length(unique(dat$site_ID)))
  )
)

cmod <- compileNimble(mod)
mcmc <- buildMCMC(mod, monitors = "lambda_derived") # Need to add monitors for derived quantity!
cMCMC <- compileNimble(mcmc)

# Mixing is much slower! Need lots of iterations
samples <- runMCMC(cMCMC, niter = 50000, nburnin = 10000, nchains = 4, thin = 1)

summary2 <- MCMCvis::MCMCsummary(samples)

summary2

bind_cols(prediction_sites, summary2) %>% 
  ggplot() +
  geom_point(aes(elevation, mean))
```



